---
title: "Principles of Indirect Treatment Comparisons"
author: "Thomas Debray"
format:
  html:
    toc: true
    number-sections: false
execute:
  message: false
  warning: false
---

```{r}
#| echo: false
#| message: false
library(dplyr)
library(knitr)
library(kableExtra)
```

This exercise helps you familiarize yourself with the principles of indirect and mixed treatment comparisons.  
You’ll use a simple, real-data-inspired example involving **Semaglutide**, **Liraglutide**, and **Placebo**, and visualize the treatment network using the `multinma` package in R.

Although the data have been slightly simplified for illustration, the reported treatment effects are based on estimates published in [*Nature Medicine* (2025)](https://www.nature.com/articles/s41591-025-03978-z) 

## Background

In this example, two clinical trials each compared an active treatment with **Placebo**.  We’d like to understand how **Semaglutide** performs relative to **Liraglutide**, even though no head-to-head trial was conducted.  We derived the following **contrast-level data** from the published trials:


| Comparison             | Effect Size | Error Variance | 95% CI (approx.)     |
|-------------------------|-------------|----------------|----------------------|
| Liraglutide vs Placebo  | 4.53        | 0.26           | (3.56, 5.55)         |
| Semaglutide vs Placebo  | 10.02       | 1.78           | (7.40, 12.63)        |

Each effect size represents the mean percentage difference in total body weight loss versus placebo. 

These effect sizes were derived from studies summarized in the Nature Medicine paper cited above.  We’ll now use them to estimate the **indirect comparison** between the two active drugs.

## Step 1. Visualizing the Treatment Network
Let’s visualize how these two studies form a (simple) network of interventions.

```{r}
#| message: false
# install.packages("multinma")  # if not already installed
library(multinma)
library(tibble)

# Contrast-level AgD with a reference row per study (y = NA on Placebo)
agd_contrast <- tribble(
  ~studyn,   ~trtn,         ~diff,        ~se_diff,
  "Study 1", "Placebo",        NA_real_,       NA_real_,
  "Study 1", "Liraglutide",     4.53,      sqrt(0.26),
  "Study 2", "Placebo",        NA_real_,       NA_real_,
  "Study 2", "Semaglutide",    10.02,     sqrt(1.78)
)

net <- set_agd_contrast(agd_contrast,
                 study = studyn,
                 trt = trtn,
                 y = diff,
                 se = se_diff)

# Visualize
plot(net, weight_edges = TRUE, weight_nodes = FALSE)
```
In this network plot, we can see

* Placebo is the common comparator
* Liraglutide and Semaglutide are connected through Placebo

## Step 2. Estimating the Indirect Comparison

When two treatments have both been compared to a common control, we can use the **consistency equation** to indirectly compare them:

$\hat{\Delta}_{A,B} = \hat{\Delta}_{A,C} - \hat{\Delta}_{B,C}$

$\text{Var}(\hat{\Delta}_{A,B}) = \text{Var}_{A,C} + \text{Var}_{B,C}$

Try it yourself:

```{r}
# Effect sizes and variances (based on Nature Medicine 2025 study)
lira <- list(est = 4.53, var = 0.26)
sema <- list(est = 10.02, var = 1.78)

# Indirect comparison: Semaglutide vs Liraglutide
est_indirect <- sema$est - lira$est
var_indirect <- sema$var + lira$var
se_indirect  <- sqrt(var_indirect)

# 95% Confidence Interval
ci <- est_indirect + qnorm(c(0.025, 0.975)) * se_indirect
```

You should obtain approximately:

```{r}
#| echo: false
df <- data.frame(
  Comparison = "Semaglutide vs Liraglutide (indirect)",
  EffectSize = est_indirect,
  SE = format(se_indirect, digits = 2, nsmall = 2),
  CI_lower = format(ci[1], digits = 2, nsmall = 2),
  CI_upper = format(ci[2], digits = 2, nsmall = 2)
)
kable(df)
```

This means that, based on indirect evidence through placebo, Semaglutide achieves about `r format(est_indirect, digits = 1, nsmall = 1)`% greater mean weight loss than Liraglutide.

We can fit the same analysis using the `multinma` package:

```{r}
#| message: false
#| warning: false
#| eval: false
db_fit_FE <- nma(net, 
                 trt_effects = "fixed",
                 link = "identity")
relative_effects(db_fit_FE, trt_ref = "Liraglutide")
```

```{r}
#| message: false
#| warning: false
db_fit_FE <- nma(net, 
                 trt_effects = "fixed",
                 link = "identity", 
                 refresh = 0)
rel_sum <- relative_effects(db_fit_FE, trt_ref = "Liraglutide")

rel_tab %>%
  select(.trtb, .trta, mean, sd, `2.5%`, `50%`, `97.5%`) %>%
  mutate(
    across(c(mean, sd, `2.5%`, `50%`, `97.5%`), ~ sprintf("%.2f", .x))
  ) %>%
  kable(
    format = "html",
    col.names = c("Treatment B", "Treatment A", "Mean", "SD", "2.5%", "Median", "97.5%"),
    align = "l"  # left alignment for text, numeric columns auto-align right
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center"
  ) %>%
  add_header_above(c(" " = 2, "Posterior summary" = 5))
```

## Step 3. Mixed Treatment Comparisons

Now that you have seen how two studies connect through a common comparator, we can extend the idea of an indirect comparison to a mixed treatment comparison.

An MTC combines both **direct** and **indirect** evidence across a connected network of trials.  
- *Direct evidence* comes from head-to-head trials comparing two treatments in the same study.  
- *Indirect evidence* arises when two treatments are compared through a shared control (e.g., Placebo).  

By "mixing" these sources, we obtain more precise estimates of relative effects across all treatments.

